---
title: "Trapping protocol"
author: "David Simons"
date: "`r format(Sys.time(), '%A %d %B %Y')`"
output:
  html_document: default
  word_document: default
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(sf)
library(tidyverse)
library(mapview)
```

## Preparation for field work

Before going to the field it's important the following are checked

  1. Adequate number of clean and functioning Sherman traps are brought. You will need at least *324* to set up the correct number so please bring some extra.
  2. Enough sample pots for rodent specimens
  3. Spare batteries for GPS devices
  4. Battery packs if possible for the electronic tablets
  5. Working portable freezer that can be stored at Panguma Hospital Lab
  6. Paper copies of the data entry form in case the pads stop working

## Location of traps

The traps will be set in grids at the same locations as previously. This means that for each grid we need to stake out the boundaries of the grid before setting the traps. We then set the traps in lines within the boundaries of the grid.

The location of each of the grids for the study sites are below. The co-ordinates are given in the same format as on the GPS devices (degrees, minutes and seconds)

``` {r map_lalehun}
lalehun_traps <- read_rds(here("data", "lalehun_traps.rds")) %>%
  st_as_sf() %>%
  distinct(geometry, .keep_all = T)

mapview(lalehun_traps)
```


